Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-13 AnalyticStructure_main.tex
continuing

When describing the Ground State of a molecule, we suppose that the potential operator, even if non-local,
is associated to real-space elements $\bra{\r} \op V \ket{\r'}$ that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite} 
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though 
perhaps in a larger region.
For a computational treatment of the problem, this property is fundamental in the 
choice of the numerical treatment; we know that basis functions that are tailored to express asymptotically vanishing 
wavefunctions are better suited for the solution of the problem.

@@

The response density operator is expressed as the first-order variation of $\dmnot=\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates\footnote{In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are described by real 
functions, when projected in the $\r$-representation.} of $\hnot$ with energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$.

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm' = \dm_{\perp} \doteq \dmnot\dm'\op Q_0 + \op Q_0\dm'\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equations \cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm']$ due to the density-dependence of $\hnot$.

@@

% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s_p^\Phi(\omega)} \;,
\ee
with:%here in this case the source term contains further addend apart from the unperturbed potential and reads:
\be
\ket{s_p^\Phi(\omega)} =  \op V\ket{f_p^\Phi(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
As a \emph{stringent} hypothesis, 
let us now assume that the operator $\op Q_0(\op V'(\omega) + \op \Phi(\omega))$ is, like $\op V$, localized, i.e. nonzero \emph{only}
on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
  %in the $\r$-representation. 
It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding 
fluctuation state. Even when the above locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $f_p(\omega,\r)$ which is \emph{delocalized} in the whole accessible space.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $f_p(\omega,\r)$ \emph{may} 
exhibit a bound-state behavior in the long range, once again provided that the above locality arguments on $\op V'$ and $\Phi$ are valid.

A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
spatial extension of the fluctuation states can be directly measured. 
The results of this procedure for a $CO$ molecule subjected to a static electric field are reported in figure \ref{co_alphaStatic}. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. The plots evidence that the typical sizes of $\ket{f_p(0)}$ are analogous to the ones of the unperturbed KS orbitals.

>>>>

When describing the Ground State of a molecule, we suppose that the potential operator, even if non-local,
can be effectively restricted on wavefunctions that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite} 
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though 
perhaps in a larger region.
For a computational treatment of the problem, this property is fundamental in the 
choice of the numerical treatment; we know that basis functions that are tailored to express asymptotically vanishing 
wavefunctions are better suited for the solution of the problem.

@@

The response density operator is expressed as the first-order variation of $\dmnot=\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates\footnote{In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are described by real 
functions, when projected in the $\r$-representation.} of $\hnot$ with (negative) energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$.

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm' = \dm_{\perp} \doteq \dmnot\dm'\op Q_0 + \op Q_0\dm'\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.

@@

% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s_p^\Phi(\omega)} \;,
\ee
with:%here in this case the source term contains further addend apart from the unperturbed potential and reads:
\be
\ket{s_p^\Phi(\omega)} =  \op V\ket{f_p^\Phi(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
As a \emph{stringent} hypothesis, 
let us now assume that the operator $\op Q_0(\op V'(\omega) + \op \Phi(\omega))$ is, like $\op V$ such that its action
can be restricted (if suitable) on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
  %in the $\r$-representation. 
It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding 
fluctuation state. Even when the above locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $f_p(\omega,\r)$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and it does not depend on the particular
choice of the operators.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $f_p(\omega,\r)$ \emph{may} 
exhibit a bound-state behavior in the long range, once again provided that the above locality arguments on $\op V'$ and $\Phi$ are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\vec F$, $\Phi_s(\r) = -\vec F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($|\vec F|$ in the above exemple) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behaviour.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}

It is then easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behaviour of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than $|\eps_p|$ this is consistent with the above observation.

% A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
% perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
% states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
% spatial extension of the fluctuation states can be directly measured. 
In Fig.~\eqref{co_alphaStatic} illustrate this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The plots evidence that the typical sizes of $\ket{f_p(0)}$ are analogous to the ones of the unperturbed KS orbitals.
